Plug-and-play dual-tree algorithm runtime analysis

### 2014-11-14 bounds-files/general-runtime-bound.tex
Rename imbalance to i_t.

\begin{equation}
\theta = \max(\log_2(\eta_r / \eta_q), 0) + i(\mathscr{T}_q) + N
\max(\log_2(\delta_q / \delta_r), 0)
\end{equation}

@@

Next, consider case \textit{(2)}.  In this situation, each query recursion
implies at least one reference recursion before another query recursion.  For
some query node $\mathscr{N}_q$, the
exact number of reference recursions before the children of $\mathscr{N}_q$ are
recursed into is bounded above by $i(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has
imbalance $0$, then it is exactly one level below its parent, and thus there is
only one reference recursion.  On the other hand, if $\mathscr{N}_q$ is many
levels below its parent, then it is possible that a reference recursion may
occur for each level in between; this is a maximum of $i(\mathscr{N}_q) + 1$.

@@

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} i(\mathscr{N}_q) + 1 = b(\mathscr{T}_q) +
O(N)
\end{equation}

@@

\begin{equation}
O\left(c^4 |R^*| \chi \psi (N + i(\mathscr{T}))\right)
\end{equation}

@@

An intuitive understanding of these bounds is best achieved by first considering
the monochromatic case (this case arises, for instance, in all-nearest-neighbor
search).  The linear dependence on $N$ arises from the fact that all query nodes
must be visited.  The dependence on the reference tree, however, is encapsulated
by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
set $R$; this value must be derived for each specific problem.  The bad
performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
also have a high cover tree imbalance $i(\mathscr{T})$; the linear dependence of
runtime on imbalance is thus sensible.

@@

The term $\theta$ captures the effect of query and reference datasets with
different widths, and has two corner cases: all extra reference recursions are of
type \textit{(1)}, and all extra reference recursions are of type \textit{(3)}.
The first case presents the best situation, because all of the reference
recursions are done before the first query node is recursed into\footnote{In
fact, re-derivation for this specific case can remove the dependence on
$i(\mathscr{T}_q)$ entirely.  But our bound, as derived, does not bound this
behavior tightly.  To do so would introduce a number of complex constants, and
thus in that respect $i(\mathscr{T}_q)$ is a more suitable bounding quantity}.
Therefore the bulk of the reference recursion work is done by with the query
tree root as the query node; this is the best case.  The worst case is the
latter, where recursion in the reference tree cannot begin until reaching the
bottom of the query tree.  Thus, no pruning can occur until reaching the leaves
of the query tree, and the recursion is essentially equivalent to a single-tree
recursion for each query point plus the added overhead of the recursion of the
query tree with one reference node.

@@

Both $\nu$ and $\kappa$ are difficult to empirically calculate and require an
entire run of the dual-tree algorithm.  On the other hand, bounding
$i(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree (assuming
the tree is already built).  Thus, not only is our bound tighter, it more
closely reflects the actual behavior of dual-tree algorithms, and the constants
which it depends upon are straightforward to calculate.

>>>>

\begin{equation}
\theta = \max(\log_2(\eta_r / \eta_q), 0) + i_t(\mathscr{T}_q) + N
\max(\log_2(\delta_q / \delta_r), 0)
\end{equation}

@@

Next, consider case \textit{(2)}.  In this situation, each query recursion
implies at least one reference recursion before another query recursion.  For
some query node $\mathscr{N}_q$, the
exact number of reference recursions before the children of $\mathscr{N}_q$ are
recursed into is bounded above by $i_t(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has
imbalance $0$, then it is exactly one level below its parent, and thus there is
only one reference recursion.  On the other hand, if $\mathscr{N}_q$ is many
levels below its parent, then it is possible that a reference recursion may
occur for each level in between; this is a maximum of $i_t(\mathscr{N}_q) + 1$.

@@

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} i_t(\mathscr{N}_q) + 1 = b(\mathscr{T}_q) +
O(N)
\end{equation}

@@

\begin{equation}
O\left(c^4 |R^*| \chi \psi (N + i_t(\mathscr{T}))\right)
\end{equation}

@@

An intuitive understanding of these bounds is best achieved by first considering
the monochromatic case (this case arises, for instance, in all-nearest-neighbor
search).  The linear dependence on $N$ arises from the fact that all query nodes
must be visited.  The dependence on the reference tree, however, is encapsulated
by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
set $R$; this value must be derived for each specific problem.  The bad
performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
also have a high cover tree imbalance $i_t(\mathscr{T})$; the linear dependence of
runtime on imbalance is thus sensible.

@@

The term $\theta$ captures the effect of query and reference datasets with
different widths, and has two corner cases: all extra reference recursions are of
type \textit{(1)}, and all extra reference recursions are of type \textit{(3)}.
The first case presents the best situation, because all of the reference
recursions are done before the first query node is recursed into\footnote{In
fact, re-derivation for this specific case can remove the dependence on
$i_t(\mathscr{T}_q)$ entirely.  But our bound, as derived, does not bound this
behavior tightly.  To do so would introduce a number of complex constants, and
thus in that respect $i_t(\mathscr{T}_q)$ is a more suitable bounding quantity}.
Therefore the bulk of the reference recursion work is done by with the query
tree root as the query node; this is the best case.  The worst case is the
latter, where recursion in the reference tree cannot begin until reaching the
bottom of the query tree.  Thus, no pruning can occur until reaching the leaves
of the query tree, and the recursion is essentially equivalent to a single-tree
recursion for each query point plus the added overhead of the recursion of the
query tree with one reference node.

@@

Both $\nu$ and $\kappa$ are difficult to empirically calculate and require an
entire run of the dual-tree algorithm.  On the other hand, bounding
$i_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
(assuming the tree is already built).  Thus, not only is our bound tighter, it
more closely reflects the actual behavior of dual-tree algorithms, and the
constants which it depends upon are straightforward to calculate.

### 2014-11-14 bounds-files/imbalance.tex
Rename imbalance to i_t.

\begin{defn}
The {\it cover node imbalance} $i(\mathscr{N}_i)$ for a cover tree node
$\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
the cumulative number of missing levels between the node and its parent
$\mathscr{N}_p$ (which has scale $s_p$).  If
the node is a leaf child (that is, $s_i = -\infty$), then number of missing
levels is defined as the difference between $s_p$ and $s_{\min} - 1$ where
$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.  If
$\mathscr{N}_i$ is the root of the tree, then the cover node imbalance is 0.
Explicitly written, this calculation is

\begin{equation}
i(\mathscr{N}_i) = \begin{dcases*}
  s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
  \max(s_p - s_{\min} - 1, \; 0) & if $\mathscr{N}_i$ is a leaf \\
  0 & if $\mathscr{N}_i$ is the root node.
  \end{dcases*}
\end{equation}
\end{defn}

@@

\begin{defn}
The {\it cover tree imbalance} $b(\mathscr{T})$ for a cover tree $\mathscr{T}$
is defined as the cumulative number of missing levels in the tree.  This can be
expressed as a function of cover node imbalances easily:

\begin{equation}
b(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} i(\mathscr{N}_i).
\end{equation}
\end{defn}

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $i(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $i(\mathscr{T}_w) \sim N (s_{\max} - s_{\min})$.
Because of this chain-like structure, each level has only one node and thus
there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning that the
imbalance is actually quadratic in $N$!

@@

{\bf TODO: Can any bound be proven on $i(\mathscr{T})$?  I haven't thought of
anything successfully yet.}

>>>>

\begin{defn}
The {\it cover node imbalance} $i_n(\mathscr{N}_i)$ for a cover tree node
$\mathscr{N}_i$ with scale $s_i$ in the cover tree $\mathscr{T}$ is defined as
the cumulative number of missing levels between the node and its parent
$\mathscr{N}_p$ (which has scale $s_p$).  If
the node is a leaf child (that is, $s_i = -\infty$), then number of missing
levels is defined as the difference between $s_p$ and $s_{\min} - 1$ where
$s_{\min}$ is the smallest scale of a non-leaf node in $\mathscr{T}$.  If
$\mathscr{N}_i$ is the root of the tree, then the cover node imbalance is 0.
Explicitly written, this calculation is

\begin{equation}
i_n(\mathscr{N}_i) = \begin{dcases*}
  s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
  \max(s_p - s_{\min} - 1, \; 0) & if $\mathscr{N}_i$ is a leaf \\
  0 & if $\mathscr{N}_i$ is the root node.
  \end{dcases*}
\end{equation}
\end{defn}

@@

\begin{defn}
The {\it cover tree imbalance} $i_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
is defined as the cumulative number of missing levels in the tree.  This can be
expressed as a function of cover node imbalances easily:

\begin{equation}
i_t(\mathscr{T}) = \sum_{\mathscr{N}_i \in \mathscr{T}} i_n(\mathscr{N}_i).
\end{equation}
\end{defn}

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $i_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $i_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that the imbalance is actually quadratic in $N$!

@@

{\bf TODO: Can any bound be proven on $i_t(\mathscr{T})$?  I haven't thought of
anything successfully yet.}
